# Air temperature - Climate inequality {#sec-airtemp3 .unnumbered}

**Goal**: Use the Imago air temperature product to explore the relationship between heat exposure and deprivation in England, moving towards regression analysis.

------------------------------------------------------------------------

## Installing packages

Here we work entirely with the panel CSV saved and the MSOA-level IMD aggregate.

::: {.panel-tabset group="language"}
### R

```{r, message=FALSE, warning=FALSE}
# Core data handling and plotting
library(tidyverse)   # Data manipulation + ggplot2

# Panel regression
library(fixest)      # Fast fixed-effects models

# Python interop (only needed if rendering both languages)
library(reticulate)
```

### Python

```{python}
# Data handling
import pandas as pd

# Plotting
import matplotlib.pyplot as plt

# Statistics
import statsmodels.formula.api as smf
import pyfixest as pf      # Fast fixed-effects models (mirrors R's fixest)
```
:::

We also redefine a couple of small helpers, so this file runs on its own:

::: {.panel-tabset group="language"}
### R

```{r}
years <- 2015:2024
```

### Python

```{python}
years = list(range(2015, 2025))

label_map = {
    "summer_max_tmp": "Summer maximum temperature (°C)",
    "extreme_hot_days_mean": "Extreme hot days (count)",
}
order = ["extreme_hot_days_mean", "summer_max_tmp"]
```
:::

# Climate inequality over time {#sec-inequality}

We now turn to the central question of this training:

> **Has the gap in heat exposure between the most and least deprived MSOAs in England widened over the decade 2015–2024?**

This question maps directly onto a literature on *unequal exposure* to environmental hazards. The key idea: deprived populations are more exposed to environmental risks, and that exposure gap can change over time as cities densify, green space is unevenly distributed, and climate change accelerates.

The strategy:

1.  Hold IMD fixed at 2025 (i.e. classify MSOAs into deprivation quintiles once),
2.  Track how mean exposure evolves in each quintile over 2015–2024,
3.  Test formally whether the gap between the most and least deprived quintiles is widening.

## Setup

We start from the panel and the MSOA-level IMD aggregate from Part 1.

::: {.panel-tabset group="language"}
### R

```{r, message=FALSE}
# Panel from Part 2
temp_panel   <- read_csv("./data/derived/temp_panel_2015_2024.csv")
msoa_imd_agg <- read_csv("./data/derived/msoa_imd_agg.csv")

# Merge in the MSOA-level health rank from Part 1, then build IMD quintiles.
# Q1 = most deprived (lowest rank), Q5 = least deprived (highest rank).
panel_imd <- temp_panel %>%
  inner_join(
    msoa_imd_agg %>% select(msoa21cd, health_rank_msoa),
    by = c("dt_zn_c" = "msoa21cd")
  ) %>%
  mutate(
    health_quintile = ntile(health_rank_msoa, 5),
    q1_dummy        = as.integer(health_quintile == 1)  # 1 = most deprived
  )

glimpse(panel_imd)
```

### Python

```{python}
# Panel from Part 2
temp_panel   = pd.read_csv("./data/derived/temp_panel_2015_2024.csv")
msoa_imd_agg = pd.read_csv("./data/derived/msoa_imd_agg.csv")

# Merge in the MSOA-level health rank from Part 1, then build IMD quintiles.
# Q1 = most deprived (lowest rank), Q5 = least deprived (highest rank).
panel_imd = (
    temp_panel
    .merge(
        msoa_imd_agg[["msoa21cd", "health_rank_msoa"]],
        how="inner", left_on="dt_zn_c", right_on="msoa21cd"
    )
    .drop(columns="msoa21cd")
)

panel_imd["health_quintile"] = pd.qcut(
    panel_imd["health_rank_msoa"], q=5, labels=False
) + 1
panel_imd["q1_dummy"] = (panel_imd["health_quintile"] == 1).astype(int)

panel_imd.info()
```
:::

::: callout-note
## Why hold IMD fixed?

The English Indices of Deprivation are released roughly every four years; the version we use here is IMD 2025. We could in principle use IMD 2019 for the early years of the panel and IMD 2025 for the later years, but two complications make this impractical for a teaching exercise: (i) IMD 2019 was published at LSOA-2011 boundaries while IMD 2025 uses LSOA-2021 boundaries, so the aggregations to MSOA are not directly comparable, and (ii) we would then be modelling change in *both* exposure and deprivation simultaneously, which is conceptually a different question. We hold deprivation fixed at IMD 2025 and let exposure vary over time. 
:::

## Headline figure: the exposure gap over time

For each year, we compute the mean of each indicator within IMD Q1 (most deprived) and within IMD Q5 (least deprived). The gap is the difference: Q1 minus Q5. Plotting the gap series over time gives the visual headline.

::: {.panel-tabset group="language"}
### R

```{r}
gap_series <- panel_imd %>%
  filter(health_quintile %in% c(1, 5)) %>%
  pivot_longer(
    cols = c(summer_max_tmp, extreme_hot_days_mean),
    names_to = "indicator", values_to = "value"
  ) %>%
  group_by(year, indicator, health_quintile) %>%
  summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from   = health_quintile,
    values_from  = mean_val,
    names_prefix = "q"
  ) %>%
  mutate(
    gap = q1 - q5,
    indicator_label = recode(indicator,
      "summer_max_tmp"        = "Summer maximum temperature (°C)",
      "extreme_hot_days_mean" = "Extreme hot days (count)"
    )
  )
```

```{r, fig.width=9, fig.height=4}
ggplot(gap_series, aes(x = year, y = gap)) +
  geom_hline(yintercept = 0, colour = "grey50", linetype = "dashed") +
  geom_line(colour = "firebrick", linewidth = 0.9) +
  geom_point(colour = "firebrick", size = 2) +
  facet_wrap(~ indicator_label, scales = "free_y") +
  scale_x_continuous(breaks = years) +
  labs(
    title    = "Exposure gap between most and least deprived MSOAs, 2015–2024",
    subtitle = "Gap = mean exposure in IMD Q1 (most deprived) − mean exposure in IMD Q5 (least deprived)",
    x = NULL, y = "Q1 − Q5",
    caption = "Source: Imago air temperature indicators; English IMD 2025."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    strip.text    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Python

```{python}
gap_long = (
    panel_imd[panel_imd["health_quintile"].isin([1, 5])]
    .melt(id_vars=["dt_zn_c", "year", "health_quintile"],
          value_vars=["summer_max_tmp", "extreme_hot_days_mean"],
          var_name="indicator", value_name="value")
    .groupby(["year", "indicator", "health_quintile"], as_index=False)
    .agg(mean_val=("value", "mean"))
)

gap_series = (
    gap_long
    .pivot_table(index=["year", "indicator"],
                 columns="health_quintile", values="mean_val")
    .reset_index()
    .rename(columns={1: "q1", 5: "q5"})
)
gap_series.columns.name = None
gap_series["gap"] = gap_series["q1"] - gap_series["q5"]
gap_series["indicator_label"] = gap_series["indicator"].map(label_map)

fig, axes = plt.subplots(1, 2, figsize=(11, 4), sharex=True)

for ax, ind in zip(axes, order):
    sub = gap_series[gap_series["indicator"] == ind]
    ax.axhline(0, color="grey", linestyle="--", linewidth=0.8)
    ax.plot(sub["year"], sub["gap"],
            color="firebrick", linewidth=1.4, marker="o")
    ax.set_title(label_map[ind], fontweight="bold")
    ax.set_xticks(years)
    ax.grid(axis="y", linestyle="--", alpha=0.4)
    ax.set_ylabel("Q1 − Q5")

fig.suptitle("Exposure gap between most and least deprived MSOAs, 2015–2024",
             fontweight="bold", y=1.02)
fig.text(0.5, -0.02,
         "Gap = mean exposure in IMD Q1 (most deprived) − mean exposure in IMD Q5 (least deprived)",
         ha="center", color="grey", fontsize=9)

plt.tight_layout()
plt.show()
```
:::

::: callout-note
## Interpretation

The gap (Q1 − Q5) is **negative in every year on both metrics**, meaning the most deprived MSOAs (IMD Q1) consistently experience *less* heat exposure than the least deprived (Q5). For summer maximum temperature the gap is fairly stable at around −1 to −1.5°C, while the extreme-hot-days gap widens sharply in hotter summers (2017, 2019, 2022 reaching −3 to −4 days).

This reflects England's geography of deprivation: the most deprived MSOAs are concentrated in the cooler north and post-industrial midlands, while the affluent south and southeast are warmer. So in the English context, the climate-justice intuition that deprived areas face greater heat exposure does not hold at the national scale — though it may well hold *within* cities, which an MSOA-level national analysis can't see.
:::

## Test 1: regress the gap on year

The simplest formal test treats the annual gap as an outcome and regresses it on year. The slope is the average annual change in the gap.

::: {.panel-tabset group="language"}
### R

```{r}
gap_smt <- gap_series %>% filter(indicator == "summer_max_tmp")
gap_ehd <- gap_series %>% filter(indicator == "extreme_hot_days_mean")

m_gap_smt <- feols(gap ~ year, data = gap_smt)
m_gap_ehd <- feols(gap ~ year, data = gap_ehd)

etable(m_gap_smt, m_gap_ehd,
       headers = c("Summer max. temp.", "Extreme hot days"),
       digits = 4)
```

### Python

```{python}
gap_smt = gap_series[gap_series["indicator"] == "summer_max_tmp"]
gap_ehd = gap_series[gap_series["indicator"] == "extreme_hot_days_mean"]

m_gap_smt = smf.ols("gap ~ year", data=gap_smt).fit()
m_gap_ehd = smf.ols("gap ~ year", data=gap_ehd).fit()

print("Summer max. temp.")
print(m_gap_smt.summary().tables[1])
print("\nExtreme hot days")
print(m_gap_ehd.summary().tables[1])
```
:::

::: callout-note

## Interpretation

Neither trend is statistically distinguishable from zero. The Q1−Q5 gap in summer maximum temperature has drifted by about **−0.007°C per year**, and the gap in extreme hot days by about **−0.003 days per year** — both with standard errors many times larger than the point estimates, and R² values essentially zero.

So the deprivation–exposure gap is **persistent but not widening or narrowing** over 2015–2024: the most deprived MSOAs remain consistently cooler than the least deprived, year after year, with no detectable trend in either direction.

Two caveats. First, with only ten yearly observations these regressions have very little statistical power — a real trend could easily be hidden by year-to-year volatility, particularly for extreme hot days. Second, a stable national-level gap can mask divergent within-region or within-city dynamics that aggregation washes out.
:::

## Test 2: panel regression with two-way fixed effects

The MSOA-level version uses the full panel. We model exposure as a function of an interaction between *being in the most deprived quintile* and a year trend, with MSOA fixed effects to absorb time-invariant local characteristics:

$$
\text{Exposure}_{it} = \alpha_i + \gamma \cdot \text{year}_t + \beta \cdot (\text{Q1}_i \times \text{year}_t) + \varepsilon_{it}
$$

The coefficient $\beta$ on the interaction is the **divergence slope** — the additional annual change in exposure for the most-deprived quintile relative to the rest. A positive, statistically significant $\beta$ means the gap is widening.

We cluster standard errors at the MSOA level to account for serial correlation within areas.

::: {.panel-tabset group="language"}
### R

```{r}
m_panel_smt <- feols(
  summer_max_tmp ~ year + i(year, q1_dummy, ref = 2015) | dt_zn_c,
  data = panel_imd
)

m_panel_ehd <- feols(
  extreme_hot_days_mean ~ year + i(year, q1_dummy, ref = 2015) | dt_zn_c,
  data = panel_imd
)

etable(m_panel_smt, m_panel_ehd,
       headers = c("Summer max. temp.", "Extreme hot days"),
       cluster = ~dt_zn_c)
```

### Python

```{python}
#m_panel_smt = pf.feols(
#    "summer_max_tmp ~ year + i(year, q1_dummy, ref=2015) | dt_zn_c",
#    data=panel_imd, vcov={"CRV1": "dt_zn_c"}
#)
#m_panel_ehd = pf.feols(
#    "extreme_hot_days_mean ~ year + i(year, q1_dummy, ref=2015) | dt_zn_c",
#    data=panel_imd, vcov={"CRV1": "dt_zn_c"}
#)
#pf.etable([m_panel_smt, m_panel_ehd],
#          model_heads=["Summer max. temp.", "Extreme hot days"])
```
:::

::: callout-note

## Main takeaway

The deprivation–heat gap is **not constant — it spikes in the hottest summers**. On average, the most deprived MSOAs are cooler than the rest of England, reflecting the country's north–south geography of deprivation. But in extreme-heat years like **2018 and 2022**, that pattern sharpens: deprived areas face **+2°C more summer heat and several additional extreme hot days** relative to their usual position. Those are precisely the years when heat causes the most harm to health.
:::

::: {.panel-tabset group="language"}
### R

```{r}
m_smt_slope <- feols(
  summer_max_tmp ~ year + year:q1_dummy | dt_zn_c,
  data = panel_imd
)

m_ehd_slope <- feols(
  extreme_hot_days_mean ~ year + year:q1_dummy | dt_zn_c,
  data = panel_imd
)

etable(m_smt_slope, m_ehd_slope,
       headers = c("Summer max. temp.", "Extreme hot days"),
       cluster = ~dt_zn_c)
```

### Python

```{python}
#m_smt_slope = pf.feols(
#    "summer_max_tmp ~ year + year:q1_dummy | dt_zn_c",
#    data=panel_imd, vcov={"CRV1": "dt_zn_c"}
#)

#m_ehd_slope = pf.feols(
#    "extreme_hot_days_mean ~ year + year:q1_dummy | dt_zn_c",
#    data=panel_imd, vcov={"CRV1": "dt_zn_c"}
#)

#pf.etable([m_smt_slope, m_ehd_slope],
#          model_heads=["Summer max. temp.", "Extreme hot days"])
```
:::

::: callout-note
## Main takeaway

Averaged across the decade, the heat-exposure gap between the most deprived MSOAs (Q1) and the rest is **slightly narrowing over time** — by about **0.0003°C and 0.0002 extreme hot days per year**. The trend is statistically significant but **substantively negligible**: at this rate it would take centuries to close even the modest existing gap. The earlier year-by-year results are the more informative reading — what matters is the *spikes* in extreme summers, not the long-run slope.
:::

## Reading the results

Three things worth pausing on:

1.  **Sign and size.** The divergence slopes are negative — the gap is narrowing — but the magnitudes (around −0.0003°C and −0.0002 days per year) are too small to matter substantively.
2.  **Statistical vs substantive significance.** With \~6,800 MSOAs × 10 years, even trivial slopes will register as highly significant. The stars aren't the interesting part.
3.  **Year-by-year vs single slope.** The year-by-year specification tells a different story: large coefficients concentrated in 2018 and 2022 rather than a smooth trend. The divergence is **episodic, tied to heatwave years**, which carries quite different policy implications than a gradual drift.

# Part 4 — Further directions

The exercise above stays in a narrow lane: one outcome (heat), one explanatory dimension (deprivation), one geography (MSOA), one country (England). Three natural extensions:

::: {.callout-tip collapse="true"}
## Option A — Comparing IMD releases

We held deprivation fixed at IMD 2025. The richer question is whether MSOAs that *became* more deprived also *became* hotter — a difference-in-differences design linking IMD 2019 to early panel years and IMD 2025 to later ones. The catch is that IMD 2019 uses LSOA-2011 boundaries and IMD 2025 uses LSOA-2021, so the LSOA-to-MSOA aggregations are not directly comparable without a rebuild via the ONS LSOA changes lookup.
:::

::: {.callout-tip collapse="true"}
## Option B — Extreme heat as a treatment

Flip the framing: treat cumulative `extreme_hot_days_mean` over 2015–2024 as the *dose* rather than the outcome. Candidate outcomes include changes in MSOA-level health deprivation (with the boundary caveats above), housing prices (Land Registry Price Paid), or migration flows (ONS internal migration estimates). Identification is the hard part — heat exposure correlates with many unmeasured local features — but this is closer to the questions regional scientists usually want to answer.
:::

::: {.callout-tip collapse="true"}
## Option C — Spatial spillovers

Our regressions treat each MSOA as independent of its neighbours; urban heat islands, green space, and deprivation all cluster across administrative boundaries. A natural extension is to test for spatial dependence (Moran's I) and estimate SLX or SAR models. In R: `spdep` and `spatialreg`; in Python: `pysal`. Particularly relevant given that the year-by-year results above suggest exposure spikes are geographically concentrated in heatwave years.
:::
