# Air temperature - Panel {#sec-airtemp2 .unnumbered}

**Goal**: Use the Imago air temperature product to explore the relationship between heat exposure and deprivation in England. We now explore a the panel data.

------------------------------------------------------------------------

## Installing packages

::: {.panel-tabset group="language"}
### R

```{r, message=FALSE, warning=FALSE}
# Core data handling and plotting
library(tidyverse)   # Data manipulation + ggplot2

# Spatial data (used to read the per-year geopackages)
library(sf)          # Simple Features

# Python interop (only needed if rendering both languages)
library(reticulate)
```

### Python

```{python}
# Standard library
import os

# Data & geospatial
import numpy as np
import pandas as pd
import geopandas as gpd

# Plotting
import matplotlib.pyplot as plt
```
:::

# Building air temperature panel {#sec-panel}

The Imago air temperature product publishes one geopackage per year, covering 2015 to 2024. In this section we:

1.  Load all ten years and stack them into a long-format panel,
2.  Keep only the two indicators we need (`summer_max_tmp` and `extreme_hot_days_mean`),
3.  Run sanity checks to confirm the panel is balanced and complete,
4.  Produce two descriptive figures showing what has happened nationally over the decade.

## Loading 2015–2024

Each yearly file follows the convention `temperature_indicators_MSOA_<year>.gpkg`. We loop over the ten years, read each file, drop the geometry (to keep the panel light), keep only the columns we need, and stack the results into a single long table.

::: {.panel-tabset group="language"}
### R

```{r, message=FALSE}
years <- 2015:2024

temp_panel <- map_dfr(years, function(yr) {
  path <- sprintf("./data/temperature_indicators_MSOA_%d.gpkg", yr)
  read_sf(path) %>%
    st_drop_geometry() %>%
    select(dt_zn_c, summer_max_tmp, extreme_hot_days_mean) %>%
    mutate(year = yr)
})

glimpse(temp_panel)
```

### Python

```{python}
years = list(range(2015, 2025))

frames = []
for yr in years:
    path = f"./data/temperature_indicators_MSOA_{yr}.gpkg"
    df = (
        gpd.read_file(path)
        .drop(columns="geometry")
        [["dt_zn_c", "summer_max_tmp", "extreme_hot_days_mean"]]
        .assign(year=yr)
    )
    frames.append(df)

temp_panel = pd.concat(frames, ignore_index=True)
temp_panel.info()
```
:::

## Panel diagnostics

Before doing anything analytical, we check that the panel behaves as expected. Three questions:

1.  How many MSOAs do we have per year?
2.  Is the panel balanced — does every MSOA appear in every year?
3.  Are there missing values in either of our two indicators?

::: {.panel-tabset group="language"}
### R

```{r}
# 1. MSOAs per year
temp_panel %>%
  count(year, name = "n_msoa")
```

```{r}
# 2. Balanced panel check — does every MSOA appear in all 10 years?
years_per_msoa <- temp_panel %>%
  count(dt_zn_c, name = "n_years")

table(years_per_msoa$n_years)
```

```{r}
# 3. Missing values per indicator per year
temp_panel %>%
  group_by(year) %>%
  summarise(
    n_total               = n(),
    miss_summer_max_tmp   = sum(is.na(summer_max_tmp)),
    miss_extreme_hot_days = sum(is.na(extreme_hot_days_mean))
  )
```

### Python

```{python}
# 1. MSOAs per year
print(temp_panel.groupby("year").size().rename("n_msoa"))

# 2. Balanced panel check — does every MSOA appear in all 10 years?
years_per_msoa = temp_panel.groupby("dt_zn_c").size().rename("n_years")
print(years_per_msoa.value_counts())

# 3. Missing values per indicator per year
print(
    temp_panel
    .groupby("year")
    .agg(
        n_total=("dt_zn_c", "size"),
        miss_summer_max_tmp=("summer_max_tmp", lambda s: s.isna().sum()),
        miss_extreme_hot_days=("extreme_hot_days_mean", lambda s: s.isna().sum()),
    )
)
```
:::

If the panel is balanced and complete, we can move on. If not, it is worth investigating before proceeding — unbalanced panels are not fatal but they do affect how some estimators behave.

## National trends over the decade

A first look: how have these two indicators evolved nationally? We compute the cross-MSOA mean and the interquartile range each year, and plot them.

::: {.panel-tabset group="language"}
### R

```{r}
national_trends <- temp_panel %>%
  pivot_longer(
    cols = c(summer_max_tmp, extreme_hot_days_mean),
    names_to = "indicator",
    values_to = "value"
  ) %>%
  group_by(year, indicator) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    q25      = quantile(value, 0.25, na.rm = TRUE),
    q75      = quantile(value, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    indicator_label = recode(indicator,
      "summer_max_tmp"        = "Summer maximum temperature (°C)",
      "extreme_hot_days_mean" = "Extreme hot days (count)"
    )
  )
```

```{r, fig.width=9, fig.height=4}
ggplot(national_trends, aes(x = year, y = mean_val)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "firebrick", alpha = 0.15) +
  geom_line(colour = "firebrick", linewidth = 0.9) +
  geom_point(colour = "firebrick", size = 1.5) +
  facet_wrap(~ indicator_label, scales = "free_y") +
  scale_x_continuous(breaks = years) +
  labs(
    title    = "National trends in heat exposure across MSOAs, 2015–2024",
    subtitle = "Solid line: cross-MSOA mean. Shaded band: interquartile range.",
    x = NULL, y = NULL,
    caption = "Source: Imago air temperature indicators."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    strip.text    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Python

```{python}
national_trends = (
    temp_panel
    .melt(id_vars=["dt_zn_c", "year"],
          value_vars=["summer_max_tmp", "extreme_hot_days_mean"],
          var_name="indicator", value_name="value")
    .groupby(["year", "indicator"], as_index=False)
    .agg(
        mean_val=("value", "mean"),
        q25=("value", lambda s: s.quantile(0.25)),
        q75=("value", lambda s: s.quantile(0.75)),
    )
)

label_map = {
    "summer_max_tmp": "Summer maximum temperature (°C)",
    "extreme_hot_days_mean": "Extreme hot days (count)",
}
national_trends["indicator_label"] = national_trends["indicator"].map(label_map)

fig, axes = plt.subplots(1, 2, figsize=(11, 4), sharex=True)
order = ["extreme_hot_days_mean", "summer_max_tmp"]

for ax, ind in zip(axes, order):
    sub = national_trends[national_trends["indicator"] == ind]
    ax.fill_between(sub["year"], sub["q25"], sub["q75"],
                    color="firebrick", alpha=0.15)
    ax.plot(sub["year"], sub["mean_val"],
            color="firebrick", linewidth=1.4, marker="o")
    ax.set_title(label_map[ind], fontweight="bold")
    ax.set_xticks(years)
    ax.grid(axis="y", linestyle="--", alpha=0.4)

fig.suptitle("National trends in heat exposure across MSOAs, 2015–2024",
             fontweight="bold", y=1.02)
fig.text(0.5, -0.02,
         "Solid line: cross-MSOA mean. Shaded band: interquartile range.",
         ha="center", color="grey", fontsize=9)

plt.tight_layout()
plt.show()
```
:::

Two things to look for in the figure: (i) is there a level trend in either indicator, and (ii) is the IQR widening, narrowing, or stable? A widening IQR would suggest *all* MSOAs are not warming together — already a hint that the inequality story may be live.

## A sample of trajectories

The national mean hides a lot. To see panel structure directly, we plot the trajectories of 50 randomly sampled MSOAs against the national mean.

::: {.panel-tabset group="language"}
### R

```{r}
set.seed(42)

sample_codes <- temp_panel %>%
  distinct(dt_zn_c) %>%
  slice_sample(n = 50) %>%
  pull(dt_zn_c)

trajectories <- temp_panel %>%
  filter(dt_zn_c %in% sample_codes) %>%
  pivot_longer(
    cols = c(summer_max_tmp, extreme_hot_days_mean),
    names_to = "indicator",
    values_to = "value"
  ) %>%
  mutate(
    indicator_label = recode(indicator,
      "summer_max_tmp"        = "Summer maximum temperature (°C)",
      "extreme_hot_days_mean" = "Extreme hot days (count)"
    )
  )

national_means <- national_trends %>%
  select(year, indicator, indicator_label, mean_val)
```

```{r, fig.width=9, fig.height=4}
ggplot() +
  geom_line(
    data = trajectories,
    aes(x = year, y = value, group = dt_zn_c),
    colour = "grey60", alpha = 0.4, linewidth = 0.3
  ) +
  geom_line(
    data = national_means,
    aes(x = year, y = mean_val),
    colour = "firebrick", linewidth = 1
  ) +
  facet_wrap(~ indicator_label, scales = "free_y") +
  scale_x_continuous(breaks = years) +
  labs(
    title    = "MSOA-level trajectories, 2015–2024",
    subtitle = "Grey lines: 50 randomly sampled MSOAs. Red line: national mean.",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey40"),
    strip.text    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Python

```{python}
rng = np.random.default_rng(42)

all_codes = temp_panel["dt_zn_c"].unique()
sample_codes = rng.choice(all_codes, size=50, replace=False)

trajectories = (
    temp_panel[temp_panel["dt_zn_c"].isin(sample_codes)]
    .melt(id_vars=["dt_zn_c", "year"],
          value_vars=["summer_max_tmp", "extreme_hot_days_mean"],
          var_name="indicator", value_name="value")
)
trajectories["indicator_label"] = trajectories["indicator"].map(label_map)

fig, axes = plt.subplots(1, 2, figsize=(11, 4), sharex=True)

for ax, ind in zip(axes, order):
    sub_t = trajectories[trajectories["indicator"] == ind]
    for code, grp in sub_t.groupby("dt_zn_c"):
        ax.plot(grp["year"], grp["value"],
                color="grey", alpha=0.4, linewidth=0.5)
    sub_n = national_trends[national_trends["indicator"] == ind]
    ax.plot(sub_n["year"], sub_n["mean_val"],
            color="firebrick", linewidth=1.6)
    ax.set_title(label_map[ind], fontweight="bold")
    ax.set_xticks(years)
    ax.grid(axis="y", linestyle="--", alpha=0.4)

fig.suptitle("MSOA-level trajectories, 2015–2024",
             fontweight="bold", y=1.02)
fig.text(0.5, -0.02,
         "Grey lines: 50 randomly sampled MSOAs. Red line: national mean.",
         ha="center", color="grey", fontsize=9)

plt.tight_layout()
plt.show()
```
:::

The trajectories make two things visible. First, individual MSOAs are noisy from year to year — single-year cross-sections (like Part 1) can be misleading. Second, the spread between MSOAs is large and the lines do not look parallel: some MSOAs may be warming faster than others. That is exactly the question Part 3 takes up.

## Save the panel

We save the assembled panel so Part 3 can pick it up without re-reading all ten geopackages.

::: {.panel-tabset group="language"}
### R

```{r}
dir.create("./data/derived", showWarnings = FALSE, recursive = TRUE)
write_csv(temp_panel, "./data/derived/temp_panel_2015_2024.csv")
```

### Python

```{python}
os.makedirs("./data/derived", exist_ok=True)
temp_panel.to_csv("./data/derived/temp_panel_2015_2024.csv", index=False)
```
:::
